---
title: "Ruby Allocation Analysis"
author: "Matt Valentine-House"
date: "07/09/2020"
output: html_document
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warnings=FALSE,
  fig.width = 14,
  fig.height = 8,
  fig.retina = 2,
  out.width = "100%"
)

r <- getOption("repos")
r["CRAN"] <- "https://www.stats.bris.ac.uk/R/"
options(repos = r)

if (!requireNamespace('tidyverse'))
  install.packages('tidyverse')

if (!requireNamespace('ggplot2'))
  install.packages('ggplot2')

if(!requireNamespace('knitr'))
  install.packages('knitr')

library(ggplot2)
library(tidyverse)
library(knitr)

options(scipen=999, dplyr.summarise.inform = FALSE)

filtered_callers = c(
    "Init_TransientHeap",
    "heap_page_allocate",
    "transient_heap_block_alloc",
    "fiber_pool_allocate_memory",
    "fiber_pool_expand",
    "Init_BareVM",
    "dln_load",
    "rb_thread_create_timer_thread"
  )

as_perc <- function(n) {
  sprintf("%1.2f%%", 100*n)
}

bar_chart <- function(data, axis, s = 'identity', title=NULL, xlab=NULL, ylab=NULL) {
  c <- ggplot(data, axis) +
  geom_bar(stat = s) +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

  if (!is.null(title))
    c = c + labs(title = title)
  if (!is.null(xlab))
    c = c + labs(x = xlab)
  if (!is.null(ylab))
    c = c + labs(y = ylab)

  c
}

group_sum_caller <- function(data) {
  data %>%
    group_by(caller) %>%
    summarise(count = n()) %>%
    arrange(desc(count))
}
group_sum_request_size_bin <- function(data) {
  data %>%
    group_by(request_size_bin) %>%
    summarise(count = n())
}

bin_grouped_percentage <- function(bin_grouped_allocations, base) {
  bin_grouped_allocations %>%
    summarise(percentage = as_perc(count / base), request_size_bin = request_size_bin) %>%
    filter(percentage >= "1.00%")
}

caller_grouped_percentage <- function(caller_grouped_allocations, base) {
  caller_grouped_allocations %>%
    summarise(percentage = as_perc(count / base), caller = caller) %>%
    filter(percentage >= "1.00%")
}

all_allocations <- read_csv('allocation_data.csv')

allocs <- all_allocations %>%
  filter(!allocator %in% c('realloc', 'reallocf'))

reallocs <- all_allocations %>%
  filter(allocator %in% c('realloc', 'reallocf'))
```

# Data Summary

```{r}
total_count <- count(all_allocations)$n
alloc_count <- count(allocs)$n
realloc_count <- count(reallocs)$n
perc_allocs <- alloc_count / total_count
perc_reallocs <- realloc_count / total_count

stats <- tibble(
  "Total" = total_count,
  "Allocation Count" = alloc_count,
  "Reallocation Count" = realloc_count,
  "Allocation Percentage" = as_perc(perc_allocs),
  "Reallocation Percentage" = as_perc(perc_reallocs)
)
kable(stats, align = 'r')
```

# Allocations

## Count of allocations binned by RVALUE size and caller

```{r message = FALSE}
filtered_allocations <- allocs %>%
  filter(!caller %in% filtered_callers)

bins <- c(-Inf, 40, seq(from = 400, to = 16320, by = 400), 16320, Inf)
labels <- c("<1", sprintf("<%d", seq(from = 10, to = 400, by = 10)), "<408", ">408")

filtered_allocations <- filtered_allocations %>%
  mutate(request_size_bin = cut(bytes_requested, breaks = bins, labels = labels))

bin_grouped_allocations <- group_sum_request_size_bin(filtered_allocations)
caller_grouped_allocations <- group_sum_caller(filtered_allocations) %>%
  filter(count > 500)

bar_chart(bin_grouped_allocations, aes(y = count, x = request_size_bin), title = "Allocations call counts grouped by request size bucket")
bar_chart(caller_grouped_allocations, aes(x = reorder(caller, count), y = count), xlab = "caller", title = "Allocation call counts, grouped by caller (where count > 500)")

bin_grouped_percentages <- bin_grouped_percentage(bin_grouped_allocations, alloc_count)
caller_grouped_percentages <- caller_grouped_percentage(caller_grouped_allocations, alloc_count)

kable(caller_grouped_percentages, align = "r", caption = "percentage of calls by caller (above 1% of all allocations)")
kable(bin_grouped_percentages, align = "r", caption = "percentage of calls by request size bin (above 1% of all allocations)")
```

## Allocations larger than a heap page (>408 RVALUES)


```{r}
total_large_allocs <- filtered_allocations %>%
  filter(request_size_bin %in% c('>408'))
```

**Total as percentage of allocations: `r as_perc(count(total_large_allocs)$n / alloc_count)`**

```{r}
large_callers_percs <- group_sum_caller(total_large_allocs) %>%
  mutate(
    percentage_of_allocs = as_perc(count / alloc_count),
    percentage_of_large_allocs = as_perc(count / count(total_large_allocs)$n))

kable(large_callers_percs, align = "r", caption = "Callers who request more than 408 RVALUES, as a percentage of all allocations")
```

```{r}
large_callers_dist <- group_sum_caller(total_large_allocs)

bar_chart(large_callers_dist, aes(y = count, x = reorder(caller, count)), xlab = "caller", title = "Large allocations by caller")
```

## Allocations smaller than 20 RVALUES

```{r}
total_small_allocs <- filtered_allocations %>%
  filter(request_size_bin %in% c('<20', '<10', '<1'))
```

**Total as percentage of allocations: `r as_perc(count(total_small_allocs)$n / alloc_count)`**

```{r}
small_callers_percs <- group_sum_caller(total_small_allocs) %>%
  mutate(
    percentage_of_allocs = as_perc(count / alloc_count),
    percentage_of_small_allocs = as_perc(count / count(total_small_allocs)$n)) %>%
  filter(percentage_of_allocs >= "1.00%")

kable(small_callers_percs, align = "r",
      caption = "Callers who request less than 20 RVALUES, as a percentage of all allocations (above 1% of all allocations)")
```

```{r}
small_callers_dist <- group_sum_caller(total_small_allocs)

qs <- quantile(small_callers_dist$count)

kable(qs, caption = "Quartile boundaries for small allocations")

small_callers_dist_upper_quartile <- small_callers_dist %>%
  filter(count > quantile(small_callers_dist$count)[3]) %>%
  arrange(desc(count))

bar_chart(small_callers_dist_upper_quartile, aes(y = count, x = reorder(caller, count)), xlab = "caller", title = "Small allocations by caller (above 50% percentile value")
```

# Reallocations

## Count of reallocations binned by RVALUE size and caller

```{r message = FALSE}
filtered_allocations <- reallocs %>%
  filter(!caller %in% filtered_callers)

bins <- c(-Inf, 40, seq(from = 400, to = 16320, by = 400), 16320, Inf)
labels <- c("<1", sprintf("<%d", seq(from = 10, to = 400, by = 10)), "<408", ">408")

filtered_allocations <- filtered_allocations %>%
  mutate(request_size_bin = cut(bytes_requested, breaks = bins, labels = labels))

bin_grouped_allocations <- group_sum_request_size_bin(filtered_allocations)
caller_grouped_allocations <- group_sum_caller(filtered_allocations)

bar_chart(bin_grouped_allocations, aes(y = count, x = request_size_bin), title = "Reallocation call counts grouped by request size bucket")
bar_chart(caller_grouped_allocations, aes(x = reorder(caller, count), y = count), xlab = "caller", title = "Rellocation call counts, grouped by caller")

bin_grouped_percentages <- bin_grouped_percentage(bin_grouped_allocations, realloc_count)
caller_grouped_percentages <- caller_grouped_percentage(caller_grouped_allocations, realloc_count)

kable(caller_grouped_percentages, align = "r", caption = "percentage of calls by caller (above 1% of all reallocations)")
kable(bin_grouped_percentages, align = "r", caption = "percentage of calls by request size bin (above 1% of all reallocations)")
```

## Reallocations larger than a heap page (>408 RVALUES)

```{r}
total_large_reallocs <- filtered_allocations %>%
  filter(request_size_bin %in% c('>408'))
```

**Total as percentage of reallocations: `r as_perc(count(total_large_reallocs)$n / realloc_count)`**

```{r}
large_callers_percs <- group_sum_caller(total_large_reallocs) %>%
  mutate(
    percentage_of_reallocs = as_perc(count / alloc_count),
    percentage_of_large_reallocs = as_perc(count / count(total_large_reallocs)$n))

kable(large_callers_percs, align = "r", caption = "Callers who request more than 408 RVALUES, as a percentage of all allocations")
```

```{r}
large_callers_dist <- group_sum_caller(total_large_reallocs)

bar_chart(large_callers_dist, aes(y = count, x = reorder(caller, count)), xlab = "caller", title = "Large allocations by caller")
```

## Reallocations smaller than 20 RVALUES

```{r}
total_small_reallocs <- filtered_allocations %>%
  filter(request_size_bin %in% c('<20', '<10', '<1'))
```

**Total as percentage of reallocations: `r as_perc(count(total_small_reallocs)$n / realloc_count)`**

```{r}
small_callers_percs <- group_sum_caller(total_small_reallocs) %>%
  mutate(
    percentage_of_reallocs = as_perc(count / alloc_count),
    percentage_of_small_reallocs = as_perc(count / count(total_small_reallocs)$n)) %>%
  filter(percentage_of_reallocs >= "1.00%")

kable(small_callers_percs, align = "r",
      caption = "Callers who request less than 20 RVALUES, as a percentage of all allocations (above 1% of all allocations)")
```

```{r}
small_callers_dist <- group_sum_caller(total_small_reallocs)

qs <- quantile(small_callers_dist$count)

kable(qs, caption = "Quartile boundaries for small allocations")

small_callers_dist_upper_quartile <- small_callers_dist %>%
  filter(count > quantile(small_callers_dist$count)[3]) %>%
  arrange(desc(count))

bar_chart(small_callers_dist_upper_quartile, aes(y = count, x = reorder(caller, count)), xlab = "caller", title = "Small allocations by caller (above 50% percentile value")

# R Versions and libraries used

```{r}
sessionInfo()
```
